The Influence of Swing States on Election Outcomes

### 2024-11-04 paper/paper.qmd

---
title: "The Coin Flip Effect"
subtitle: "How Swing States Transform Elections Into Games of Chance"
author: 
  - Tara Chakkithara
  - Sakura Hu
  - Luka Tosic
thanks: "Code and data are available at: https://github.com/LukaTosic09/Forecasting-2024-Election"
date: today
date-format: long
abstract: "This paper uses the poll of polls method to develop a bayesian hierarcical model to predict the outcome of the 2024 U.S. election. Based on polling data collected up to October 19, 2024, our analysis forecasts that Kamala Harris is likely to win the election. Given the complexities of the electoral college system, this study specifically focuses on predicting the electoral college vote rather than the popular vote."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

library(tidyverse)
library(janitor)
library(lubridate)
library(broom)
library(modelsummary)
library(rstanarm)
library(splines)

@@

# Data {#sec-data}

# Data Overview (missing graphs and tables)

@@

```{r}
#| include: false
#| warning: false
#| message: false
#| echo: false

election_data <- read.csv("../data/raw_data/president_polls.csv")
real_data <- election_data %>% filter(hypothetical=="false", answer == "Harris")

real_data <- real_data %>% mutate(national = ifelse(is.na(state), TRUE, FALSE))
quality_data <- real_data %>% filter(sample_size >= 1000, numeric_grade>=3.0)

#still having trouble with dates
#real_data$end_date <- as.Date(real_data$end_date, format="%d/%m/%y")
#real_data <- real_data %>% mutate(end_date = mdy(end_date))# %>% filter(!is.na(end_date))
#fixed_date <- ymd("2024-25-10")
#real_data$recency <- as.numeric(difftime(fixed_date, real_data$end_date, units = "days"))
real_data %>% mutate(end_date = mdy(end_date))

```{r}
ggplot(real_data, aes(x=numeric_grade, y=pollscore)) + geom_point() + geom_smooth(formula = y ~ x, method = "lm",se = FALSE)+labs(title = "what's the relationship even?")
```

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-predictor vs response plots
#| height: 20
#| width: 6

ggplot(real_data, aes(x=pollster, y=pct)) + geom_point() + geom_smooth(formula = y ~ x , method = "lm", color = "black", se = FALSE) + labs(title = "pollster vs pct") #+ theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

ggplot(real_data, aes(x=state, y=pct)) + geom_point() + geom_smooth(formula = y ~ x , method = "lm", color = "black", se = FALSE) + labs(title = "state vs pct") +  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

ggplot(real_data, aes(x=methodology, y=pct)) + geom_point() + geom_smooth(formula = y ~ x , method = "lm", color = "black", se = FALSE) + labs(title = "method vs pct") #+  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

ggplot(real_data, aes(x=methodology)) + geom_bar() + labs(title="methodology occurences") +   theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

ggplot(real_data, aes(x=partisan, y=pct)) + geom_point() + geom_smooth(formula = y ~ x , method = "lm", color = "black", se = FALSE) + labs(title = "partisan vs pct") +  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

```{r}
ggplot(real_data, aes(x=sponsor_candidate_party, y=pct)) + geom_point() + geom_smooth(formula = y ~ x , method = "lm", color = "black", se = FALSE) + labs(title = "sponsor party vs pct") +  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

ggplot(real_data, aes(x=transparency_score, y=numeric_grade)) + geom_point() + geom_smooth(formula = y ~ x , method = "lm", color = "black", se = FALSE) + labs(title = "trans vs poll") +  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

```{r}
test_model <- lm(pct ~ end_date + state, data = quality_data)
test_model

>>>>

---
title: "The Influence of Swing States on Election Outcomes"
author: 
  - Tara Chakkithara
  - Sakura Hu
  - Luka Tosic
thanks: "Code and data are available at: https://github.com/LukaTosic09/Forecasting-2024-Election"
date: today
date-format: long
abstract: "This paper employs a Bayesian hierarchical model to forecast the outcome of the 2024 U.S. election. Based on polling data collected from July 21 to October 19, 2024, our analysis indicates that Kamala Harris is the candidate most likely to win. Given the complexities of the Electoral College voting system and significant political polarization, election outcomes appear very close this year. This paper aims to quantify this closeness and provide insights into voter behavior."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

library(tidyverse)
library(janitor)
library(lubridate)
library(broom)
library(modelsummary)
library(rstanarm)
library(splines)
library(ggthemes)
library(sf)
library(dplyr)
library(readr)
```

@@

# Data {#sec-data}
## Data Acquisition
Data for this paper is sourced from the Project 538 Presidential General Election polls dataset @cite538. This dataset includes approximately 16,000 observations of polls conducted from January 2023 to October 16, 2024.
Each poll in the dataset has an associated pollscore, sample size, start date, pct (percentage of votes), candidate name, and state. There are other features include time of publication, candidate sponsorship, etc, however for this analysis we will focus on the key veriables highlighted.

## Dataset Features
### Pollscore

```{r}
#| warning: false
#| message: false
#| echo: false

election_data <- read.csv("../data/raw_data/president_polls.csv")

real_data <- election_data %>% select(pollscore)

ggplot(real_data, aes(x = pollscore)) +
  geom_density(fill = "#4B8BBE", alpha = 0.6) + 
  labs(title = "Pollscore Distribution",
       x = "Pollscore",
       y = "Density") +
  theme_minimal() + 
  theme(
    plot.title = element_text(hjust = 0.5, size = 10, face = "plain"), 
    axis.title.x = element_text(size = 10, face = "plain"),             
    axis.title.y = element_text(size = 10, face = "plain")               
  )
```

### Sample Size

```{r}
#| echo: false
#| warning: false
#| message: false

data <- read_csv('../data/raw_data/president_polls.csv')

data <- data |>
  select(pollscore, transparency_score)

data <- data |>
  group_by(pollscore) |>
  summarise(avg_transparency_score = mean(transparency_score, na.rm = TRUE), .groups = "drop")

ggplot(data, aes(x = pollscore, y = avg_transparency_score)) +
  geom_point(color = 'saddlebrown', size = 3) +
  labs(
    x = 'Pollscore',
    y = 'Average Transparency Score',
    title = 'Average Transparency Score by Pollscore'
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 10), 
    axis.title.x = element_text(hjust = 0.46, size = 10, margin = margin(t = 10)),  
    axis.title.y = element_text(size = 10, margin = margin(r = 10)),
    axis.text = element_text(size = 9)
  )

```{r}
#| echo: false
#| warning: false
#| message: false

election_data <- read.csv("../data/raw_data/president_polls.csv")

real_data <- election_data %>% select(sample_size)

ggplot(real_data, aes(x = sample_size)) +
  geom_density(fill = "#4B8BBE", alpha = 0.6) + 
  labs(title = "Sample Size",
       x = "Sample Size",
       y = "Density") +
  xlim(0, 10000) +
  theme_minimal() + 
  theme(
    plot.title = element_text(hjust = 0.5, size = 10, face = "plain"), 
    axis.title.x = element_text(size = 10, face = "plain"),             
    axis.title.y = element_text(size = 10, face = "plain")               
  )
```

```{r}
#| echo: false
#| warning: false
#| message: false

swing_states <- c('Arizona', 'Michigan', 'Georgia', 'Nevada', 
                  'North Carolina', 'Pennsylvania', 'Wisconsin')

map_data <- st_read('../data/raw_data/s_05mr24/s_05mr24.shp', quiet = TRUE)

data <- read_csv('../data/raw_data/president_polls.csv')

biden_drop_date <- as.Date("2024-07-21")

data <- data |>
  mutate(start_date = as.Date(sub("(..)$", "20\\1", start_date), format = "%m/%d/%Y")) |>
  filter(start_date < biden_drop_date) |>
  filter(pollscore < -1) |>
  select(state, pct, candidate_name) |>
  filter(candidate_name %in% c("Donald Trump", "Joe Biden")) |>  
  filter(state %in% swing_states) |>
  group_by(state, candidate_name) |>  
  summarise(avg_pct = mean(pct, na.rm = TRUE), .groups = "drop")

map_data <- map_data |> 
  rename(state = NAME)

wide_data <- data |>
  pivot_wider(names_from = candidate_name, values_from = avg_pct) |>  
  rename(Trump = `Donald Trump`, Biden = `Joe Biden`)

merged_data <- map_data |>
  left_join(wide_data, by = "state")

merged_data <- merged_data |>
  mutate(color_category = case_when(
    Trump > Biden ~ "Donald Trump",
    Biden > Trump ~ "Joe Biden",
    TRUE ~ "Tie" 
  ))

ggplot(data = merged_data) +
  geom_sf(aes(fill = color_category), color = "white", size = 3, show.legend = FALSE) + 
  scale_fill_manual(values = c("Donald Trump" = "red", "Joe Biden" = "blue", "Tie" = "gray"), 
                    name = "Winner") + 
  labs(
    title = "Voter Sentiment During Biden's Presidential Campaign",
    x = NULL,
    y = NULL
  ) +
  coord_sf(xlim = c(-125, -66), ylim = c(24, 50), expand = FALSE) +  
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 10, face = "plain"),
    legend.direction = "horizontal", 
    legend.title = element_text(hjust = 0.5)
  )
```

```{r}
#| echo: false
#| warning: false
#| message: false

swing_states <- c('Arizona', 'Michigan', 'Georgia', 'Nevada', 
                  'North Carolina', 'Pennsylvania', 'Wisconsin')

map_data <- st_read('../data/raw_data/s_05mr24/s_05mr24.shp', quiet = TRUE)

data <- read_csv('../data/raw_data/president_polls.csv')

biden_drop_date <- as.Date("2024-07-21")

data <- data |>
  mutate(start_date = as.Date(sub("(..)$", "20\\1", start_date), format = "%m/%d/%Y")) |>
  filter(start_date > biden_drop_date) |>
  filter(pollscore < -1) |>
  select(state, pct, candidate_name) |>
  filter(candidate_name %in% c("Donald Trump", "Kamala Harris")) |>  
  filter(state %in% swing_states) |>
  group_by(state, candidate_name) |>  
  summarise(avg_pct = mean(pct, na.rm = TRUE), .groups = "drop")

map_data <- map_data |> 
  rename(state = NAME)

wide_data <- data |>
  pivot_wider(names_from = candidate_name, values_from = avg_pct) |>  
  rename(Trump = `Donald Trump`, Harris = `Kamala Harris`)

merged_data <- map_data |>
  left_join(wide_data, by = "state")

merged_data <- merged_data |>
  mutate(color_category = case_when(
    Trump > Harris ~ "Donald Trump",
    Harris > Trump ~ "Kamala Harris",
    TRUE ~ "Tie" 
  ))

ggplot(data = merged_data) +
  geom_sf(aes(fill = color_category), color = "white", size = 3, show.legend = FALSE) + 
  scale_fill_manual(values = c("Donald Trump" = "red", "Kamala Harris" = "blue", "Tie" = "gray"), 
                    name = "Winner") + 
  labs(
    title = "Voter Sentiment after Biden Drops out of the Presidential Race",
    x = NULL,
    y = NULL
  ) +
  coord_sf(xlim = c(-125, -66), ylim = c(24, 50), expand = FALSE) +  
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 10, face = "plain"),
    legend.direction = "horizontal", 
    legend.title = element_text(hjust = 0.5)
  )

```{r}
#| echo: false
#| warning: false
#| message: false

data <- read_csv('../data/raw_data/president_polls.csv')

biden_drop_date <- as.Date("2024-07-21")

data <- data |>
  mutate(start_date = as.Date(sub("(..)$", "20\\1", start_date), format = "%m/%d/%Y")) |>
  mutate(end_date = as.Date(sub("(..)$", "20\\1", end_date), format = "%m/%d/%Y")) |>
  filter(start_date > biden_drop_date) |>
  filter(is.na(state)) |>
  filter(candidate_name %in% c("Kamala Harris", "Donald Trump")) |>  
  group_by(end_date, candidate_name) |>  
  summarise(avg_pct = mean(pct, na.rm = TRUE), .groups = "drop")

wide_data <- data |> 
  pivot_wider(names_from = candidate_name, values_from = avg_pct, 
              values_fill = list(avg_pct = NA))

ggplot(data = wide_data, aes(x = end_date)) +
  geom_line(aes(y = `Kamala Harris`, color = "Kamala Harris"), size = 1) + 
  geom_line(aes(y = `Donald Trump`, color = "Donald Trump"), size = 1) + 
  labs(
    title = "Time Series of Average Polling Percentages",
    x = "Date",
    y = "Average Percentage",
    color = "Candidate"
  ) +
  scale_color_manual(values = c("Kamala Harris" = "blue", "Donald Trump" = "red")) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "top"
  )

# Data Overview

@@

@@
